---
title: "Plotting Frequencies"
subtitle: "SDS 192: Introduction to Data Science"
author: "Professor Lindsay Poirier"
format: 
  revealjs:
    scrollable: true
editor: source
---

## For Today

-   Fill out CATME Survey
-   Histograms
-   Bar plots
-   Multivariate Frequencies

------------------------------------------------------------------------

# Frequency plots (histograms and barplots) involve *counting* the values in a variable.

------------------------------------------------------------------------

## Let's create the following data frame to motivate today's lecture.

```{r}
library(tidyverse)
counties <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_counties.csv")
route_prefixes <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_route_pre.csv")
maintenance <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_maintenance.csv")
kinds <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_kind.csv")

nbi_ma <- read.delim("https://www.fhwa.dot.gov/bridge/nbi/2024/delimited/MA24.txt", sep = ",") |>
  left_join(counties) |>
  left_join(route_prefixes) |>
  left_join(maintenance) |>
  left_join(kinds) |>
  filter(SERVICE_ON_042A == 1) |>
  select(STRUCTURE_NUMBER_008, COUNTY_CODE_003_L, ROUTE_PREFIX_005B_L, MAINTENANCE_021_L, YEAR_BUILT_027, ADT_029, STRUCTURE_KIND_043A_L, STRUCTURAL_EVAL_067, BRIDGE_IMP_COST_094) |>
  mutate(STRUCTURE_KIND_043A_L = 
           case_when(
             STRUCTURE_KIND_043A_L == "Concrete continuous" ~ "Concrete",
             STRUCTURE_KIND_043A_L == "Steel continuous" ~ "Steel",
             STRUCTURE_KIND_043A_L == "Prestressed concrete continuous" ~ "Prestressed concrete",
             TRUE ~ STRUCTURE_KIND_043A_L)) |>
  mutate(BRIDGE_IMP_COST_094 = BRIDGE_IMP_COST_094 * 1000)

nbi_hampshire <- nbi_ma |> filter(COUNTY_CODE_003_L == "Hampshire")

rm(counties, kinds, maintenance, route_prefixes)
```

------------------------------------------------------------------------

## Histogram

::: columns
::: {.column width="50%"}

-   Visualizes *distribution* of a **numeric** variable
    -   What are maximum and minimum values?
    -   How spread out are the values?
    -   What is the center of the values.
:::

::: {.column width="50%"}

```{r}
#| echo: false
df <- data.frame(x = rnorm(100, mean = 50, sd = 20))
ggplot(df, aes(x = x, y="")) + geom_point(size = 3, alpha = 0.5) +
  theme(text = element_text(size = 20))
ggplot(df, aes(x = x)) +
  geom_histogram(color = "white", binwidth = 10)  +
  theme(text = element_text(size = 20))
```

:::
:::

------------------------------------------------------------------------

## Histogram

::: columns
::: {.column width="50%"}

1. Create bins for numbers, each with the same range of values [i.e. 0-10, >10-20, >20-30, and so on]
    - Converts the linear scale to a categorical scale
2. Count the numbers in each bin
3. Set the height of a bar for that bin to the count 

:::

::: {.column width="50%"}

```{r}
#| echo: false
df <- data.frame(x = rnorm(100, mean = 50, sd = 20))
ggplot(df, aes(x = x, y="")) + geom_point(size = 3, alpha = 0.5) +
  theme(text = element_text(size = 20))
ggplot(df, aes(x = x)) +
  geom_histogram(color = "white", binwidth = 10) +
  theme(text = element_text(size = 20))
```

:::
:::

------------------------------------------------------------------------

# Learning Check: What's the scale of the x-axis on the previous plot? What's the scale of the y-axis?

------------------------------------------------------------------------

## Bar Plot

::: columns
::: {.column width="50%"}

-   Visualizes *counts* of a **categorical** variable
    -   Which value appears the most?
    -   Which appears the least?
    -   How evenly distributed are the counts?

:::

::: {.column width="50%"}

```{r}
#| echo: false
df <- data.frame(x =c("a", "b", "c", "a", "c", "a", "a", "b", "c", "a", "b", "c"))
df$x
ggplot(df, aes(x = x)) +
  geom_bar(color = "white") +
  theme(text = element_text(size = 20))
```

:::
:::

------------------------------------------------------------------------


## Bar Plot

::: columns
::: {.column width="50%"}

1. Determine the unique values and places them on the x-axis
2. Count the number of times each value appears
3. Set the height of a bar for that category to the count 

:::

::: {.column width="50%"}

```{r}
#| echo: false
df <- data.frame(x =c("a", "b", "c", "a", "c", "a", "a", "b", "c", "a", "b", "c"))
df$x
ggplot(df, aes(x = x)) +
  geom_bar(color = "white") +
  theme(text = element_text(size = 20))
```

:::
:::

------------------------------------------------------------------------

# Learning Check: What's the scale of the x-axis on the previous plot? What's the scale of the y-axis?

------------------------------------------------------------------------


## Distribution in MA Bridge's Years Built


```{r}
ggplot(nbi_ma, aes(x = YEAR_BUILT_027)) +
  geom_histogram()  +
  theme(text = element_text(size = 20))
```


------------------------------------------------------------------------


## Bidwidth vs. Bins

::: columns
::: {.column width="50%"}

-   Binwidth indicates the width of the buckets we'd like to categorize our data into.
-   Bins indicates the number of bins to create. 
-   We choose one or the other when creating histograms. 


:::

::: {.column width="50%"}

```{r}
ggplot(nbi_ma, 
       aes(x = YEAR_BUILT_027)) +
  geom_histogram(binwidth = 10, 
                 color = "white")
```

:::
:::


------------------------------------------------------------------------


## How would we describe this plot?

```{r}
ggplot(nbi_ma, aes(x = YEAR_BUILT_027)) +
  geom_histogram(binwidth = 10, color = "white") +
  labs(title = "Distribution in the Years MA Bridges were Built, 2021", 
       x = "Year Built",
       y = "Number of Bridges")
```



------------------------------------------------------------------------

## Faceting a Histogram


```{r}
ggplot(nbi_ma, aes(x = YEAR_BUILT_027)) +
  geom_histogram(binwidth = 10, color = "white") +
  labs(title = "Distribution in the Years MA Bridges were Built, 2021", 
       x = "Year Built",
       y = "Number of Bridges") +
  facet_wrap(vars(ROUTE_PREFIX_005B_L))
```


------------------------------------------------------------------------

## Frequency of Structure Kinds

```{r}
ggplot(nbi_ma, aes(x = STRUCTURE_KIND_043A_L)) +
  geom_bar()
```


------------------------------------------------------------------------

## Labels for this Plot


```{r}
ggplot(nbi_ma, aes(x = STRUCTURE_KIND_043A_L)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Frequency of Different Kinds of MA Bridge Structures, 2021", 
       x = "Structure Kind",
       y = "Number of Bridges")
```


------------------------------------------------------------------------

## Stacked Bar Plot

```{r}
ggplot(nbi_ma, aes(x = STRUCTURE_KIND_043A_L, fill = ROUTE_PREFIX_005B_L)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Frequency of Different Kinds of MA Bridge Structures, 2021", 
       x = "Structure Kind",
       y = "Number of Bridges",
       fill = "Route Prefix") +
  scale_fill_brewer(palette = 'Set3')
```





------------------------------------------------------------------------

## Learning Check: Why not this?

```{r}
#| echo: false
ggplot(nbi_ma, aes(x = STRUCTURE_KIND_043A_L, fill = ROUTE_PREFIX_005B_L)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Frequency of Different Kinds of MA Bridge Structures, 2021", 
       x = "Structure Kind",
       y = "Number of Bridges",
       fill = "Route Prefix") +
  scale_fill_brewer(palette = 'Blues')
```


------------------------------------------------------------------------

## Dodging 


```{r}
ggplot(nbi_ma, aes(x = STRUCTURE_KIND_043A_L, fill = ROUTE_PREFIX_005B_L)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Frequency of Different Kinds of MA Bridge Structures, 2021", 
       x = "Structure Kind",
       y = "Number of Bridges",
       fill = "Route Prefix") +
  scale_fill_brewer(palette = 'Set3')
```



------------------------------------------------------------------------

## Converting to Percentages

```{r}

ggplot(nbi_ma, aes(x = STRUCTURE_KIND_043A_L, fill = ROUTE_PREFIX_005B_L)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Frequency of Different Kinds of MA Bridge Structures, 2021", 
       x = "Structure Kind",
       y = "Number of Bridges",
       fill = "Route Prefix") +
  scale_fill_brewer(palette = 'Set3')
```