---
title: "Functions"
subtitle: "SDS 192: Introduction to Data Science"
author: "Professor Lindsay Poirier"
format: 
  revealjs:
    scrollable: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 5
editor: source
execute:
  echo: true
  messages: false
  warnings: false
---


# In your small groups, design a formula to quantitatively measure a student's level of self-care this week of the semester. 

------------------------------------------------------------------------

## Why write functions?

```{r echo=FALSE}
library(tidyverse)
pit <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/main/website/data/pit_2015_2020.csv") |>
  pivot_longer(-State, names_to = "Measure", values_to = "Count") |>
  separate(Measure, into = c("Shel_Unshel", "Demographic"), sep = " - ") |>
  separate(Demographic, into = c("Demographic", "Year"), sep = ", ") |>
  filter(Demographic %in% c("Male", "Female", "Transgender")) |>
  filter(State != "Total")
```

```{r}
pit %>%
  filter(State == "AK" & Year == 2015) |>
  summarize(Total = sum(Count))

pit %>%
  filter(State == "AL" & Year == 2015) |>
  summarize(Total = sum(Count))

pit %>%
  filter(State == "AR" & Year == 2015) |>
  summarize(Total = sum(Count))
```

------------------------------------------------------------------------

## Why write functions?

-   Reduces the amount of code to write
-   Lowers the chances of errors in code
-   Makes it easier to update code as requirements of a project change
-   Supports reproducibility

------------------------------------------------------------------------

## Functions

-   Statements organized to perform a specific task
-   Take arguments as *inputs*
    -   Arguments can be required or optional
-   Return a value or set of values as *outputs*
    -   Default value is output of last line of function if not specified
  
------------------------------------------------------------------------

## User-defined Functions

::: panel-tabset

### Basic format

```{r eval=FALSE}
function_name <- function(arg1, arg2) {
  
  x <- #Some code goes here referencing arg1 and arg2
  
  return(x)
}
```

### Example

```{r}
calculate_difference <- function(arg1, arg2) {
  
  x <- arg1 - arg2
  
  return(x)
}
```

:::

------------------------------------------------------------------------

## Calling Functions


-   This is the same way you'd call functions built-in to `R`


::: panel-tabset

### Basic format

```{r eval = FALSE}
function_name(value_for_arg1, value_for_arg2)
```

### Example

```{r}
calculate_difference(7, 3)
```

:::

------------------------------------------------------------------------

## Function Best Practices

-   Function names should communicate what a function will do
-   Function names should often be verbs
-   Argument names should be nouns
-   Function should include comments that explain what the function does


------------------------------------------------------------------------

## User-defined Functions

-   `state_name` is a required argument here. There is no default value for the argument in the function.

```{r}
calculate_state_total <- function(state_name) {
  
  x <- pit %>%
    filter(State == state_name & Year == 2015) |>
    summarize(Total = sum(Count))
  
  return(x)
}
```

```{r}
calculate_state_total(state_name = "AL")
```


------------------------------------------------------------------------

## Making Arguments Optional

-   Because we provide a default value for year, it is optional in the function call. 

::: panel-tabset

### Setting Variable Defaults

```{r}
calculate_state_total <- function(state_name, year = 2020) {
  
  x <- pit %>%
    filter(State == state_name & Year == year) |>
    summarize(Total = sum(Count))
  
  return(x)
}
```

```{r}
calculate_state_total("AL")
```

### Overriding Variable Defaults

```{r}
calculate_state_total <- function(state_name, year = 2020) {
  
  x <- pit %>%
    filter(State == state_name & Year == year) |>
    summarize(Total = sum(Count))
  
  return(x)
}
```

```{r}
calculate_state_total("AL", 2019)
```

:::

------------------------------------------------------------------------


## Naming Arguments

-   Naming helps to differentiate between arguments. Order matters if arguments aren't named!

```{r}
calculate_state_total <- function(state_name, year = 2020) {
  
  x <- pit %>%
    filter(State == state_name & Year == year) |>
    summarize(Total = sum(Count))
  
  return(x)
}
```

```{r}
calculate_state_total(state_name = "AL", year = 2019)
```

------------------------------------------------------------------------


## Return Statements

-   If your function does not include a return statement, it will return the last statement the code evaluated.
-   If it includes a return statement, it will return the value passed to `return()`.

```{r}
calculate_state_total <- function(state_name, year = 2020) {
  
  x <- pit %>%
    filter(State == state_name & Year == year) |>
    summarize(Total = sum(Count))
  
  sum(2, 3)
}
```

```{r}
calculate_state_total(state_name = "AL", year = 2019)
```

------------------------------------------------------------------------


## If Else Statements

-   You can execute code only when certain conditions are met using if-else statements. 


```{r}
calculate_state_total <- function(state_name, year = 2020) {
  
  x <- pit %>%
    filter(State == state_name & Year == year) |>
    summarize(Total = sum(Count))
  
  if (x >= 3000) {
    return("Total is greater than or equal to 3000")
  }
  else if (x >= 2000) {
    return("Total is greater than or equal to 3000 but less than 3000.")
  }
  else {
    return("Total is less than 2000.")
  }
}
```

```{r}
calculate_state_total(state_name = "AL", year = 2019)
```