---
title: "`ggplot()`"
subtitle: "SDS 192: Introduction to Data Science"
author: "Professor Lindsay Poirier"
format: 
  revealjs:
    scrollable: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 5
editor: source
execute:
  echo: true
  messages: false
  warnings: false
---


## For Today

-   Visualizations Exercise
-   Introduction to `ggplot()`
-   Class Activity

------------------------------------------------------------------------


## Let's create the following data frame to motivate today's lecture. 

> This dataset comes from [Pioneer Valley Data](https://pioneervalleydata.org/data-download-page/) and documents estimates of population characteristics for each municipality in the Pioneer Valley. 

```{r}
library(tidyverse)
pioneer_valley_census_data <- read_csv("https://raw.githubusercontent.com/SDS-192-Intro/public-website-fall-25/refs/heads/main/data/pioneer_valley_census_2022.csv")

pioneer_valley_census_data_dictionary <- pioneer_valley_census_data |> 
  select(VAR_CAT, VAR, VAR_NAME) |>
  distinct()

pioneer_valley_census_data <- pioneer_valley_census_data |>
  select(-VAR_CAT, -VAR) |>
  pivot_wider(names_from = VAR_NAME,
              values_from = VALUE) |>
  filter(LEVEL_CD_NAME != "Region")
  
hampshire_census_data <- pioneer_valley_census_data %>% 
  filter(COUNTY == "Hampshire")
```

------------------------------------------------------------------------

## `ggplot`

-   Most plots we create in this course will rely on package called `ggplot2`
-   `ggplot2` is included in the Tidyverse, which you installed in SDS 100
-   Load `ggplot2` in your environment. 

```{r}
library(ggplot2)
```

------------------------------------------------------------------------

## Anatomy of the `ggplot()` function

-   `ggplot()` takes two arguments:
    -   data: the dataset used to produce the plot (in a data frame)
    -   mapping: the variables from the dataset we want mapped onto visual cues
        -   mappings are defined in a function called `aes()` (short for aesthetics)
        -   in Cartesian plots, we must supply the variables/columns that will appear on the axes (via `x = ` and `y = `)

------------------------------------------------------------------------

## Anatomy of the `ggplot()` function

```{r, fig.height=5, fig.retina=2, fig.width=10}
ggplot(data = hampshire_census_data, 
       aes(x = COMMUNITY, 
           y = 	CEN_CHILD_POV))
```

------------------------------------------------------------------------

# Learning check: What's the scale of the x-axis in the plot you just created? What's the scale of the y-axis?

------------------------------------------------------------------------

## Where's the data?

-   In previous plot, we told `R` *what* variables to plot, but we didn't indicate *how* to plot them.
-   To do this, we need to add a *geom function* to our `ggplot` call. Examples:
    -   Bar plots: `geom_bar()`
    -   Scatterplots: `geom_point()`
-   Appended to function call with a `+` sign

------------------------------------------------------------------------

## Adding a geom function

```{r, fig.height=5, fig.retina=2, fig.width=10}
ggplot(data = hampshire_census_data, 
       aes(x = COMMUNITY, 
           y = 	CEN_CHILD_POV)) +
  geom_col()
```

------------------------------------------------------------------------

# Learning check: What variables are mapped on to what visual cues in this plot?

------------------------------------------------------------------------

## Styling Plots: Flipping Coordinates

```{r, fig.height=5, fig.retina=2, fig.width=10}
ggplot(data = hampshire_census_data, 
       aes(x = COMMUNITY, 
           y = CEN_CHILD_POV)) +
  geom_col() +
  coord_flip() # Flipping the x and y coordinates here makes the labels more legible.
```

------------------------------------------------------------------------

# Learning check: How's the data-to-ink ratio on this plot?

------------------------------------------------------------------------

## Styling Plots: Changing the Theme

```{r, fig.height=5, fig.retina=2, fig.width=10}
ggplot(data = hampshire_census_data, 
       aes(x = COMMUNITY, 
           y = 	CEN_CHILD_POV)) +
  geom_col() +
  coord_flip() + # Flipping the x and y coordinates here makes the labels more legible. 
  theme_minimal()
```

------------------------------------------------------------------------

# Learning check: What context needs to be added to this plot?

------------------------------------------------------------------------

## Styling Plots: Adding Labels

```{r, fig.height=5, fig.retina=2, fig.width=10}
ggplot(data = hampshire_census_data, 
       aes(x = COMMUNITY, 
           y = CEN_CHILD_POV)) +
  geom_col() +
  coord_flip() + # Flipping the x and y coordinates here makes the labels more legible. 
  theme_minimal() +
  labs(title = "Hampshire County Child Poverty, 2022", 
       x = "Municipality in Hampshire County, MA", 
       y = "Child Poverty Rate")
```

------------------------------------------------------------------------

## Styling Plots: Adjusting the Scale

```{r, fig.height=5, fig.retina=2, fig.width=10}
# Adjust the Scale
ggplot(data = pioneer_valley_census_data, 
       aes(x = COUNTY,y = CEN_SINGPARHOU)) +
  geom_point() +
  coord_flip() +
  scale_y_log10() +
  labs(title = "Single Parent Households in Pioneer Valley, MA Municipalities, 2018", x = "County", y = "Number of Households")
```

------------------------------------------------------------------------

## Aeshetics vs. Attributes

-   We can adjust the way the data appears on plots in two ways:
    -   According to a variable: 
        -   This must be done inside of the `aes()` function
    -   In a fixed way:
        -   This must be done outside of the `aes()` function

------------------------------------------------------------------------

## Adjusting Data on Plots via Aeshetics

> We add visual cues to the plot in the `aes()` call

```{r, fig.height=5, fig.retina=2, fig.width=10}
# Add visual cue for size

ggplot(data = pioneer_valley_census_data, 
       aes(x = COUNTY,y = CEN_SINGPARHOU, size = CEN_HOUSEHOLDS)) +
  geom_point() +
  coord_flip() +
  scale_y_log10() +
  labs(title = "Single Parent Households in Pioneer Valley, MA Municipalities, 2018", x = "County", y = "Number of Single Parent Households", size = "Total Households")
```

------------------------------------------------------------------------

## Adjusting Data on Plots via Attributes

```{r, fig.height=5, fig.retina=2, fig.width=10}
# Add visual cue for size and attribute for transparency
ggplot(data = pioneer_valley_census_data, 
       aes(x = COUNTY,y = CEN_SINGPARHOU, size = CEN_HOUSEHOLDS)) +
  geom_point(alpha = 0.5) +
  coord_flip() +
  scale_y_log10() +
  labs(title = "Single Parent Households in Pioneer Valley, MA Municipalities, 2018", x = "County", y = "Number of Single Parent Households", size = "Total Households")
```

------------------------------------------------------------------------

## Do I really have to memorize all of these stylistic functions?!

No. There are cheatsheets. The `ggplot2()` cheatsheet is linked [here](https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf).