---
title: "Iteration"
subtitle: "SDS 192: Introduction to Data Science"
author: "Professor Lindsay Poirier"
format: 
  revealjs:
    scrollable: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 5
editor: source
execute:
  echo: true
  messages: false
  warnings: false
---

## For Loops

```{r}
#| echo: false
library(tidyverse)
df <- data.frame(
  name = c("obs1", "obs2", "obs3", "obs1"),
  var_a = c(2, 3, 4, 5),
  var_b = c(4, 7, 2, 1),
  var_c = c(4, 9, 3, 2)
)

df
```


```{r}
for (i in df$var_a) {
  print(i + 1)
}
```

------------------------------------------------------------------------

## For Loops

```{r}
#| echo: false
df
```


```{r}
for (i in df |> select(var_a:var_c)) {
  print(sum(i))
}
```

------------------------------------------------------------------------

## `across()` 

-   Applies a function across multiple columns in a data frame
-   Takes as arguments: the columns to perform the function across and the function name

::: panel-tabset

### Example 1

```{r}
df |>
  summarize(across(var_a:var_c, sum))
```

### Example 2

```{r}
df |>
  summarize(across(contains("var"), mean))
```

### Example 3

```{r}
df |>
  mutate(across(where(is.numeric), as.character))
```

:::
  
------------------------------------------------------------------------

## `purrr` package

-   Included in `tidyverse`
-   Package for working with functions and vectors
-   Provides a family of `map()` functions
-   `map()` functions allow us to apply a function to each element of a list or vector

------------------------------------------------------------------------

## Single Column Data Frames vs. Vectors

- To extract a column from a data frame, we use `pull()`

::: panel-tabset

### Data Frame

```{r}
df |> select(var_a)
```

### Vector (`$`)

```{r}
df$var_a
```

### Vector (`pull()`)

```{r}
df |> select(var_a) |> pull()
```

:::

------------------------------------------------------------------------

`map()`

- Applies a function to each element in a vector

::: panel-tabset

### Approach 1

```{r}
add_five <- function(x){
  x + 5
}

map(df$var_a, add_five)
```

### Approach 2

```{r}
add_five <- function(x){
  x + 5
}

a_vec <- df |> select(var_a) |> pull()
map(a_vec, add_five)
```

:::

------------------------------------------------------------------------

## Setting Names

-   `set_names()` sets the names of elements in a vector

```{r}
add_five <- function(x){
  x + 5
}

map(df$var_a, add_five) |>
  set_names(df$name)
```


------------------------------------------------------------------------

## Family of Map Functions

::: panel-tabset

### `map_int()`

-   Returns a numeric vector

```{r}
df |>
  select(var_a:var_c) |>
  map_int(is.numeric)
```

### `map_chr()`

-   Returns a character vector

```{r}
df |>
  select(var_a:var_c) |>
  map_chr(is.numeric)
```

### `map_lgl()`

-   Returns a logical vector

```{r}
df |>
  select(var_a:var_c) |>
  map_lgl(is.numeric)
```

:::

------------------------------------------------------------------------

## Returning a Data Frame

::: panel-tabset

### `map()`

-   Returns a list

```{r}
create_total_col <- function(x){
  df |>
    filter(name == x) |>
    mutate(total = var_a + var_b + var_c)
    
}

map(unique(df$name), create_total_col)
```

### `map_df()`

-   Returns a data frame (binding rows of list)

```{r}
create_total_col <- function(x){
  df |>
    filter(name == x) |>
    mutate(total = var_a + var_b + var_c)
    
}

map_df(unique(df$name), create_total_col)
```

:::

------------------------------------------------------------------------

## Iterating Over Multiple Vectors

```{r}
add_two_vectors <- function(x, y){
  x + y
}

map2(df$var_a, df$var_b, add_two_vectors)
```