---
title: Advanced Data Wrangling
editor: source
format:
  html:
    embed-resources: true
---

```{r unnamed-chunk-1}
#| warning: false
#| message: false
#| document: show

library(tidyverse)

sqf_url <- "https://www1.nyc.gov/assets/nypd/downloads/zip/analysis_and_planning/stop-question-frisk/sqf-2011-csv.zip"
temp <- tempfile()
download.file(sqf_url, temp)
sqf_zip <- unzip(temp, "2011.csv")
sqf_2011 <- read.csv(sqf_zip, stringsAsFactors = FALSE) 
sqf_2011_race_cat <- read.csv("https://raw.githubusercontent.com/lindsaypoirier/STS-101/master/Data/SQF/sqf_race_categories.csv", stringsAsFactors = FALSE) 
rm(sqf_url)
rm(temp)
rm(sqf_zip)
```
###### Question 1
Select all columns relating the crime suspected.

```{r unnamed-chunk-2}
#Code here
```
###### Question 2
Select all of the first 5 columns, except the 4th column. 

```{r unnamed-chunk-4}
# Code Here
```
###### Question 3
Send me the distinct values in the crime suspected column.

```{r unnamed-chunk-6}
# Code Here
```
###### Question 4
In which rows where a suspect was frisked and searched was either contraband or a pistol found on the suspect? Add a line of code below to determine this. 

```{r unnamed-chunk-8}
sqf_2011 |>
  select(frisked, searched, contrabn, pistol) |> 
  head(5)
```
###### Question 5
Send me all of the rows where the crime suspected contains "FEL."

```{r unnamed-chunk-10}
# Code Here
```
###### Question 6
The NYPD uses the number 999 to indicate a missing age. In the code below, add a line to convert 999 to NA, using `mutate()`. 

```{r unnamed-chunk-12}
# Code Here

sqf_2011 |>
  filter(age > 100) |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 100)
```
###### Question 7
Add a line to the code below to create a new column called `innocent` that is set to "Y" if both `arstmade` and `sumissue` are "N", and otherwise set to "N".

```{r unnamed-chunk-14}
sqf_2011 |>
  select(pct, ser_num, arstmade, sumissue) #|>
  #summarize(num_innocent = sum(innocent == "Y"))
```
###### Question 8
In the following code, the percent total stops is calculated out the total stops for each city. How can I adjust the code if I wanted to calculate percent total stops out of the total stops in the entire dataset.

```{r unnamed-chunk-16}
sqf_2011 |>
  group_by(city, race) |>
  summarize(count = n()) |>
  mutate(total_stops = sum(count)) |>
  mutate(percent_total_stops = count/total_stops)
```
###### Question 9
In which precinct were individuals of each race stopped the most?

```{r unnamed-chunk-18}
# Code Here
```
