---
title: 'Lab 7: Tidying Data'
format:
  html:
    embed-resources: true
editor: source
knitr:
  opts_chunk:
    message: false
    warning: false
---

```{r import}
#| document: show
library(tidyverse)
pit <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/main/website/data/pit_2015_2020.csv") 

#Note for future reference how adding the following could remove unreliable counts:
# mutate_at(vars(2:97), funs(case_when(. < 20 ~ as.numeric(NA), TRUE ~ .)))

gender <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/main/website/data/gender_state_2015_2020.csv")
race <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/main/website/data/race_state_2015_2020.csv")
```
###### Question 1

Write code to pivot longer columns 2 through 117 (i.e. all columns except the first state column) in `pit`. Store the names in column called "Measure" and store the values in a column called "Value." Store the resulting data frame in in `pit_pivoted`.


```{r unnamed-chunk-2}
# Uncomment below and write code to pivot dataset. Store the results in pit_pivoted

# pit_pivoted <- pit |> pivot_longer(_____)
```
###### Question 2

Write code to separate `Measure` into two columns. One column should be called `Shel_Unshel`, and the second should be called `Demographic`. Note what symbol separates these two pieces of data: space dash space (i.e. " - "). You will need to specify this in the `sep` argument for `separate()`. Store the resulting data frame in in `pit_separate_1`.


```{r unnamed-chunk-5}
# Uncomment below and complete the code to separate the Measure column. Store the results in pit_separate_1

# pit_separate_1 <- pit_pivoted |> separate(_____)
```
###### Question 3

Write code to separate `Demographic` into two columns. One column should be called `Demographic`, and the second should be called `Year`. Note what symbol separates these two pieces of data: comma space (i.e. ", "). You will need to specify this in the `sep` argument for `separate()`. Store the resulting data frame in in `pit_separate_2`.


```{r unnamed-chunk-8}
# Uncomment below and complete the code to separate the Demographic column. Store the results in pit_separate_2

# pit_separate_2 <- pit_separate_1 |> separate(_____)
```
###### Question 4

Remove the string "Total " from the `Shel_Unshel` column. To do this you should use the `str_replace()` function. Replace the string "Total" with the empty string (i.e. "") to remove these characters. Store the results in `pit_str_cleaned`.


```{r unnamed-chunk-11}
# Uncomment below and complete the code to replace the string "Total " in Shel_Unshel with an empty string. Store the results in pit_str_cleaned

# pit_str_cleaned <- pit_separate_2 |> mutate(Shel_Unshel = _______)
```
###### Question 5

Create two new data frames - one for `pit_by_gender` and one for `pit_by_race`. To do this you want to extract the rows with values in Demographic `%in%` the following vector of values: `c("Female", "Male", "Transgender")`, and store the result in `pit_by_gender`. Then you want to extract the rows with values in Demographic `%in%` the following vector of values: 

```
c("Black or African American", 
  "Asian", 
  "American Indian or Alaska Native", 
  "Native Hawaiian or Other Pacific Islander", 
  "White", 
  "Multiple Races")
```

and store the result in `pit_by_race`.


```{r unnamed-chunk-14}
# Uncomment below to create two new tables.

# pit_by_gender <- pit_str_cleaned |>
# pit_by_race <- pit_str_cleaned |>
```
```{r unnamed-chunk-17}
#| document: show

rm(pit, pit_pivoted, pit_separate_1, pit_separate_2, pit_str_cleaned)
```
###### Question 6

Create two new tidy data frames - `gender_final` and `race_final`. For each, you should do this in 3 steps:

Step 1: Pivot the columns `estimate_2015:moe_2020` longer, storing the column names in "Measure" and the values in "Values".
Step 2: Separate the values in Measure into two columns: "Measure", "Year". Note that these are separated by an underscore ("_").
Step 3: Pivot the Measure column wider, taking the values from the Values column. 

```{r unnamed-chunk-21}
# Fill in the blanks to clean up these data frames. 

# Step 1
#gender_pivoted_longer <- gender |> pivot_longer(_____)

# Step 2
#gender_separated <- gender_pivoted_longer |> separate(_____)

# Step 3
#gender_final <- gender_separated |> pivot_wider(_____)

# Step 1
#race_pivoted_longer <- race |> pivot_longer(_____)

# Step 2
#race_separated <- race_pivoted_longer |> separate(_____)

# Step 3
#race_final <- race_separated |> pivot_wider(_____)
```
```{r unnamed-chunk-24}
#| document: show
rm(gender, gender_pivoted_longer, gender_separated, race, race_pivoted_longer, race_separated)
```
```{r unnamed-chunk-26}
#| document: show
pit_by_gender <- 
  pit_by_gender |>
  mutate(StateName = state.name[match(State, state.abb)])
  
pit_by_race <- 
  pit_by_race |>
  mutate(StateName = state.name[match(State, state.abb)])
```
###### Question 7

Join `pit_by_gender` on the left to `gender_final` on the right and store the results in `pit_gender`. Join `pit_by_race` on the left to `race_final` on the right and store the results in `pit_race`. In both case, you will be joining by three variables: State, Demographic, and Year. We can set the left key variables to the right key variables by setting the `by` argument to this in the join: `c("StateName" = "NAME", "Demographic" = "variable", "Year" = "Year")`


```{r unnamed-chunk-27}
# Uncomment below and write code to join pit_by_gender to gender_final and pit_by_race to race_final. Store the results in pit_gender and pit_race respectively.

# pit_gender <- pit_by_gender |>
# pit_race <- pit_by_race |>
```
###### Question 8

Use a data wrangling verb to create a new column in both `pit_gender` and `pit_race` that calculates the rate of homelessness per 10,000 population. Set that column name to `homeless_rate`. Hint: When creating that column, you'll need to divide the homeless count by the population estimate and then multiply by 10000. Store the resulting data frames in `pit_gender_rates` and `pit_race_rates`.


```{r unnamed-chunk-30}
# Uncomment below and write code to create a new column for homelessness_rate in both pit_gender and pit_race

# pit_gender_rates <- pit_gender |>
# pit_race_rates <- pit_race |>
```
```{r unnamed-chunk-33}
#| document: show
options(scipen=999)

pit_gender_rates |>
  filter(State == "FL") |>
  ggplot(aes(x = Year, 
             y = homeless_rate, 
             col = Demographic, 
             group = Demographic)) +
  geom_line() +
  facet_wrap(vars(Shel_Unshel)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Point in Time Homeless Rates in FL, 2015-2020",
       x = "Year", 
       y = "Homeless per 10,000 Population",
       col = "Gender")

pit_race_rates |>
  filter(State == "FL") |>
  ggplot(aes(x = Year, 
             y = homeless_rate, 
             col = Demographic, 
             group = Demographic)) +
  geom_line() +
  facet_wrap(vars(Shel_Unshel)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 3, byrow = TRUE)) +
  labs(title = "Point in Time Homeless Rates in FL, 2015-2020",
       x = "Year", 
       y = "Homeless per 10,000 Population",
       col = "Race")

```
