---
title: R Fundamentals
editor: source
format:
  html:
    embed-resources: true
---

```{r unnamed-chunk-1}
#| message: false
#| warning: false
#| document: show

library(tidyverse)
nyc_urban_ranger <- 
  read_csv("https://data.cityofnewyork.us/api/views/fuhs-xmg2/rows.csv",
           name_repair = make.names) |>
  mutate_if(is.character, na_if, c("")) |>
  mutate_if(is.character, na_if, c("N/A"))
```
###### Question 1

1. Apply a function below to determine the number of rows in the data frame.


```{r unnamed-chunk-2}
# Write code here
```
###### Question 2
2. Apply a function below to determine the data frame's column names.

```{r unnamed-chunk-3}
# Write code here
```
###### Question 3

3. Apply a function below to determine the unique values in the `Species.Status` column.

```{r unnamed-chunk-4}
# Write code here
```
###### Question 4
4. Apply a function below to determine the sum of the `X..of.Animals` column. Be sure to account for `NA` values. What does this represent?

```{r unnamed-chunk-5}
# Write code here
```
###### Question 5
5. Apply a function below to determine the number of missing values in the `Animal Condition` column.


```{r unnamed-chunk-6}
# Write code here
```
