---
title: "Advanced Data Wrangling"
editor: source
---

```{r}
#| warning: false
#| message: false
#| echo: false
#| document: show

library(tidyverse)

sqf_url <- "https://www1.nyc.gov/assets/nypd/downloads/zip/analysis_and_planning/stop-question-frisk/sqf-2011-csv.zip"
temp <- tempfile()
download.file(sqf_url, temp)
sqf_zip <- unzip(temp, "2011.csv")
sqf_2011 <- read.csv(sqf_zip, stringsAsFactors = FALSE) 
sqf_2011_race_cat <- read.csv("https://raw.githubusercontent.com/lindsaypoirier/STS-101/master/Data/SQF/sqf_race_categories.csv", stringsAsFactors = FALSE) 
rm(sqf_url)
rm(temp)
rm(sqf_zip)
```

###### Question

::: question
Select all columns relating the crime suspected.
:::

```{r}
#Code here
```

```{r}
#| document: solutions
#| echo: false

sqf_2011 |>
  select(starts_with("cs")) |> 
  head(5)
```

###### Question

::: question
Select all of the first 5 columns, except the 4th column. 
:::

```{r}
# Code Here
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  select(1:5, -4) |> 
  head(5)
```

###### Question

::: question
Send me the distinct values in the crime suspected column.
:::

```{r}
# Code Here
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  select(crimsusp) |>
  distinct() |> 
  head(5)
```

###### Question

::: question
In which rows where a suspect was frisked and searched was either contraband or a pistol found on the suspect? Add a line of code below to determine this. 
:::

```{r}
sqf_2011 |>
  select(frisked, searched, contrabn, pistol) |> 
  head(5)
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  filter(frisked == "Y" & searched == "Y" & (contrabn == "Y" | pistol == "Y")) |>
  select(frisked, searched, contrabn, pistol) |> 
  head(5)
```

###### Question

::: question
Send me all of the rows where the crime suspected contains "FEL."
:::

```{r}
# Code Here
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  filter(str_detect(crimsusp, "FEL")) |>
  select(pct, ser_num, crimsusp) |> 
  head(5)
```

###### Question

::: question
The NYPD uses the number 999 to indicate a missing age. In the code below, add a line to convert 999 to NA, using `mutate()`. 
:::

```{r}
# Code Here

sqf_2011 |>
  filter(age > 100) |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 100)
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  mutate(age = na_if(age, 999)) |>
  filter(age > 100) |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 100)
```

###### Question

::: question
Add a line to the code below to create a new column called `innocent` that is set to "Y" if both `arstmade` and `sumissue` are "N", and otherwise set to "N".
:::

```{r}
#sqf_2011 |>
  #select(pct, ser_num, arstmade, sumissue) #|>
  #summarize(num_innocent = sum(innocent == "Y"))
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  mutate(innocent = case_when(arstmade == "Y" & sumissue == "Y" ~ "N", 
                              TRUE ~ "Y")) |>
  select(pct, ser_num, arstmade, sumissue, innocent) |>
  summarize(num_innocent = sum(innocent == "Y"))
  
```

###### Question

::: question
In the following code, the percent total stops is calculated out the total stops for each city. How can I adjust the code if I wanted to calculate percent total stops out of the total stops in the entire dataset.
:::

```{r}
sqf_2011 |>
  group_by(city, race) |>
  summarize(count = n()) |>
  mutate(total_stops = sum(count)) |>
  mutate(percent_total_stops = count/total_stops)
```

```{r}
#| document: solutions
#| echo: false


sqf_2011 |>
  group_by(city, race) |>
  summarize(count = n()) |>
  ungroup() |>
  mutate(total_stops = sum(count)) |>
  mutate(percent_total_stops = count/total_stops)
```

###### Question

::: question
In which precinct were individuals of each race stopped the most?
:::

```{r}
# Code Here
```

