---
title: "Day 6: Frequency`"
format: html
editor: source
---


```{r}
#| message: false
#| warning: false
#| document: show

library(tidyverse)
counties <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_counties.csv")
route_prefixes <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_route_pre.csv")
maintenance <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_maintenance.csv")
kinds <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_kind.csv")

nbi_ma <- read.delim("https://www.fhwa.dot.gov/bridge/nbi/2022/delimited/MA22.txt", sep = ",") |>
  left_join(counties) |>
  left_join(route_prefixes) |>
  left_join(maintenance) |>
  left_join(kinds) |>
  filter(SERVICE_ON_042A == 1) |>
  select(STRUCTURE_NUMBER_008, COUNTY_CODE_003_L, ROUTE_PREFIX_005B_L, MAINTENANCE_021_L, YEAR_BUILT_027, ADT_029, STRUCTURE_KIND_043A_L, STRUCTURAL_EVAL_067, BRIDGE_IMP_COST_094) |>
  mutate(STRUCTURE_KIND_043A_L = 
           case_when(
             STRUCTURE_KIND_043A_L == "Concrete continuous" ~ "Concrete",
             STRUCTURE_KIND_043A_L == "Steel continuous" ~ "Steel",
             STRUCTURE_KIND_043A_L == "Prestressed concrete continuous" ~ "Prestressed concrete",
             TRUE ~ STRUCTURE_KIND_043A_L)) |>
  mutate(BRIDGE_IMP_COST_094 = BRIDGE_IMP_COST_094 * 1000)

rm(counties, kinds, maintenance, route_prefixes)
```

###### Question

::: question

Check column names and values.

:::

```{r}

#Check the column names for nbi_ma
```

###### Question

::: question
Create a plot to visualize the distribution of structural evaluations for bridges in MA. What does the height of the bar represent?
:::

```{r}
# Plot Here
```

###### Question

::: question
Facet the previous plot by county
:::

```{r}
# Plot Here
```

###### Question

::: question
Create a plot to visualize the frequency of bridges in each county. What does the height of the bar represent?
:::

```{r}
# Plot Here
```

###### Question

::: question
Color the previous plot by route prefix, and set the position to dodge. If you finish early, experiment with adjust the plot's palette. (Refer to `ggplot()` cheatsheet and lecture slides for help.)
:::

```{r}
# Plot Here
```
