---
title: "Point Mapping"
editor: visual
---

```{r}
#| warning: false
#| message: false
#| echo: false
#| document: show

library(tidyverse)
library(leaflet)
library(sf)

counties <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_counties.csv")
maintenance <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_maintenance.csv")
kinds <- read_csv("https://raw.githubusercontent.com/sds-192-intro-fall22/sds-192-public-website-quarto/a8b64e3070ca2543b904d4d92780b09e6062ced6/website/data/nbi_kind.csv")

nbi_ma <- read.delim("https://www.fhwa.dot.gov/bridge/nbi/2024/delimited/MA24.txt", sep = ",") |>
  left_join(counties) |>
  left_join(maintenance) |>
  left_join(kinds) |>
  filter(SERVICE_ON_042A == 1) |>
  select(STRUCTURE_NUMBER_008, COUNTY_CODE_003_L, MAINTENANCE_021_L, ADT_029, STRUCTURE_KIND_043A_L, STRUCTURAL_EVAL_067, LAT_016, LONG_017) |>
  mutate(STRUCTURE_KIND_043A_L = 
           case_when(
             STRUCTURE_KIND_043A_L == "Concrete continuous" ~ "Concrete",
             STRUCTURE_KIND_043A_L == "Steel continuous" ~ "Steel",
             STRUCTURE_KIND_043A_L == "Prestressed concrete continuous" ~ "Prestressed concrete",
             TRUE ~ STRUCTURE_KIND_043A_L)) |>
  mutate(LAT_016_DEG = as.numeric(substr(LAT_016, 1, 2)),
         LAT_016_MIN = as.numeric(substr(LAT_016, 3, 4)),
         LAT_016_SEC = as.numeric(substr(LAT_016, 5, 8)),
         LONG_017_DEG = as.numeric(substr(LONG_017, 1, 2)),
         LONG_017_MIN = as.numeric(substr(LONG_017, 3, 4)),
         LONG_017_SEC = as.numeric(substr(LONG_017, 5, 8)),
         LAT = LAT_016_DEG + (LAT_016_MIN/60) + (LAT_016_SEC/360000),
         LONG = (LONG_017_DEG + (LONG_017_MIN/60) + (LONG_017_SEC/360000)) * -1)


rm(counties, kinds, maintenance)
```

###### Question

::: question
Convert `nbi_ma` to an `sf` object. The CRS is 4326.
:::

```{r}
# Code Here
```

```{r}
#| document: solutions
#| echo: false

nbi_ma <- nbi_ma  |>
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326)
```

###### Question

::: question
Initiate a basemap of MA and add "CartoDB.Positron" provider tiles. You can use coordinates: 42.4072 (lat) -71.3824 (long) and set the zoom to 7.
:::

```{r}
#Code here
```

```{r}
#| document: solutions
#| echo: false

leaflet() |>
  setView(lat = 42.4072, lng = -71.3824, zoom = 7) |>
  addProviderTiles("CartoDB.Positron") 
```


###### Question

::: question
Add Circle Markers of bridges to the map. 
:::

```{r}
# Code Here
```

```{r}
#| document: solutions
#| echo: false

leaflet() |>
  setView(lat = 42.4072, lng = -71.3824, zoom = 7) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = nbi_ma)
```

###### Question

::: question
Adjust the `radius`, `stroke`, and `fillOpacity`.  
:::

```{r}
# Code Here
```

```{r}
#| document: solutions
#| echo: false

leaflet() |>
  setView(lat = 42.4072, lng = -71.3824, zoom = 7) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = nbi_ma,
                   radius = 2,
                   fillOpacity = 0.8,
                   stroke = FALSE)
```

###### Question

::: question
Set the `fillColor` to `ADT_029`, and swap out the different palettes. I've given you code to add a legend. Be sure to update the palette in the legend when swapping out palettes. Which palette do you think is most appropriate?
:::

```{r}

pal_numeric <- colorNumeric(palette = "YlOrRd", 
                    domain = nbi_ma$ADT_029)
  
pal_bin <- colorBin(palette = "YlOrRd", 
                    domain = nbi_ma$ADT_029, 
                    bins = 4)

pal_quant <- colorQuantile(palette = "YlOrRd", 
                    domain = nbi_ma$ADT_029, 
                    n = 4)


# Code Here

# addLegend(data = nbi_ma,
#             title = "Annual Daily Traffic of Bridges, 2024", 
#             pal = pal_numeric, 
#             values = ~ADT_029)
  
```

```{r}
#| document: solutions
#| echo: false

leaflet() |>
  setView(lat = 42.4072, lng = -71.3824, zoom = 7) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = nbi_ma,
                   radius = 2,
                   fillOpacity = 0.8,
                   stroke = FALSE,
                   fillColor = ~pal_bin(ADT_029)) |>
  addLegend(data = nbi_ma,
            title = "Annual Daily Traffic of Bridges, 2024", 
            pal = pal_bin, 
            values = ~ADT_029)
```

###### Question

::: question
Set the `fillColor` to `STRUCTURAL_EVAL_067`, and swap out the different palettes. I've given you code to add a legend. Be sure to update the palette in the legend when swapping out palettes. Note how I reversed the scale. Which palette do you think is most appropriate?
:::

```{r}

pal_numeric <- colorNumeric(palette = "YlOrRd", 
                    domain = nbi_ma$STRUCTURAL_EVAL_067,
                    reverse = TRUE)
  
pal_bin <- colorBin(palette = "YlOrRd", 
                    domain = nbi_ma$STRUCTURAL_EVAL_067, 
                    bins = 4,
                    reverse = TRUE)

pal_quant <- colorQuantile(palette = "YlOrRd", 
                    domain = nbi_ma$STRUCTURAL_EVAL_067, 
                    n = 4,
                    reverse = TRUE)


# Code Here

# addLegend(data = nbi_ma,
#             title = "Structural Evaluation of Bridges, 2024", 
#             pal = pal_numeric, 
#             values = ~STRUCTURAL_EVAL_067)
  
```

```{r}
#| document: solutions
#| echo: false

leaflet() |>
  setView(lat = 42.4072, lng = -71.3824, zoom = 7) |>
  addProviderTiles("CartoDB.Positron") |>
  addCircleMarkers(data = nbi_ma,
                   radius = 2,
                   fillOpacity = 0.8,
                   stroke = FALSE,
                   fillColor = ~pal_numeric(STRUCTURAL_EVAL_067)) |>
  addLegend(data = nbi_ma,
            title = "Structural Evaluation of Bridges, 2024", 
            pal = pal_numeric, 
            values = ~STRUCTURAL_EVAL_067)
```

